Wireless Forensics Principles and Investigations

### 2021-12-03 paper.tex
Fixed some missing citations

\section{Mobile Forensics}
\subsection{Introduction to Mobile Forensics}
Mobile device forensics is a subcategory of digital forensics.  It involves the recovery of digital evidence or data from any mobile device.  These mobile devices include mobile phones, smart phones, tablets, personal digital assistants (PDAs), and Global Positioning Systems (GPS).  Another type of mobile device is wearable technology, for example an Apple watch or Fitbit fitness tracker.  These types of mobile devices can store several types of private information.  This information consists of call history, contact list, incoming and outgoing text messages, pictures, videos, audio files and voicemail messages. With the usage of smart phones, the data set also comprises internet browsing history, previous Wi-Fi connections, calendars, analytical information, geolocation data, and documents.  Some people have been known to save notes with account user credentials, passwords, and passcodes in their mobile devices as well.  All this information could be useful in an investigation.  Given the wide variety of mobile devices, there are different ways each device version must be handled.  The differentiation of the data on the device depends on the cell phone generation, the device brand, and the version of the specific model.  As a part of digital forensics investigations, there are specific tools that should be used to ensure the integrity of the mobile devices during transport from one place to the next.  Additionally, there are various data collecting techniques used for data analysis and preservation.

\subsection{Progression of the Cellular Networks}
Cellular networks and technologies have changed massively since their arrival in the late 1970s.  The first-generation mobile networks started in Japan in 1979 before other countries like the US and United Kingdom.  This technology was known as the advanced mobile phone system (AMPS).  The AMPS used frequency division multiple modulation and only had a capacity of 30 kilohertz and 2.4 kbps of speed.  The 1g network only allowed voice calls and had limited protection against hackers.  The second-generation network started around 1991.  This system was based on digital signals from the global system for mobile communications (GSM).  The 2g network offered bandwidths of 30 kHz to 200 kHz.  The users could also send SMS/MMS messages up to 64 kb.  Soon after, 2.5g was established and enabled data rates up to 144 kbps.  These rates allowed for browsing the web and sending   and receiving email messages. In the year 2000, the third-generation network was introduced in Europe as UMTS and CDMA in the US.  The introduction of 3G became more about social connections and supported higher speeds.  The 3G upgrade also enabled video calls, sharing files, watching television, and more surfing the internet and now for playing online games.  The fourth-generation network era brought about the usage of smartphones and other handheld devices.  These devices have speeds from 10 Mbps and 1 Gbps.  Fourth generation was the first cellular network generation to use long term evolution (LTE), which offers users better voice quality, less buffering time, instant messaging and social media, faster downloads, and quality streaming.    The 4G cellular network is also the first IP based mobile network with the technology to accommodate the quality of service and requirements including broadband access, multimedia messaging services (MMS), video chat, mobile TV, HDTV content, and digital video broadcasting.  As the 4g network tries to keep up with the demand of augmented reality and development of autonomous vehicles, the 5g revolution has begun.  The 5th generation network is being created to enhance the growing rate of new mobile applications and standards.  The output of 5G devices is 1000 times faster than 4G at 10 Gbps and will support the number of IoT (internet of things) sensors and devices.

@@

Data collecting and preservation for mobile device forensics is getting more complicated as the devices advance.  Mobile devices are used for both personal and professional purposes, so the data contained reflects that.  To recover the digital evidence and other relevant data, you must follow the guidelines for preservation and examination.  All the methods and techniques must be forensically sound, meaning all actions and processes used must respect international guidelines and avoid the alteration of the original devices.  For mobile device forensics, digital forensics in general as well, evidence should always be preserved in process in a manner that allows it to be used in a court of law.  When moving mobile devices from the scene, there are two things to consider: lock activation and network/cellular connection.  Preventing network connection is important to ensure that the mobile device is not changed remotely.  The mobile device can be isolated by placing the device on airplane mode to disable Wi-Fi and hotspots.  It is also important to make sure the device has a full charge to avoid shutdown.  During storage and transportation, Faraday bags are a great tool to use.  A simple Faraday bag will isolate the device from receiving radio signals while keeping the device charged with an emergency battery.  All original cords and chargers should be removed from the scene and secluded away from the device.  Faraday cases are made from silver, copper, or nickel.  They have RoHS (Restriction of Hazardous Substances) double layer conductors, which are best for blocking EMF (electromagnetic fields) signals.  Another tool that could be used for isolation during transportation is a jammer.  A jammer sends radio waves with the same frequency as mobile phones, which causes interference between communications and paralyzes every phone in the range of the jammer.  In some countries jammers are illegal and are only allowed to be used by law enforcement.  There are many data collecting techniques.  Your plan of execution as a digital forensics professional, must include an outline of actions taken in sequence.  This is necessary because the process may have to be duplicated at another time.  Mobile forensics data collecting methods can be non-invasive and invasive.  As a good rule of thumb, always start with non-invasive techniques because “they tend to endanger the device is integrity to a lesser degree” (Kostadinov, 2019).  Types of noninvasive mobile forensics data extraction methods include manual extraction, logical extraction, hex dump, and JTAG.  When arriving at the crime scene some devices may already be unlocked.  If this is the case, then manual extraction is possible.  Manual extraction is when you browse through the mobile device’s data by using the touch screen or keypad.  Ideally during manual extraction, you would document and take photos of the device while you're browsing.  Logical extraction involves connecting the mobile device to a forensically sound workstation.  The connection is established by using USB cords, Bluetooth, or RJ-45 cables.  Command requests for data retrieval from the computer to the mobile device.  With this method, data could be added to the mobile device altering the integrity.  The hexdump process is performed when the mobile device is connected to a forensically sound workstation, then using a bootloader a set of instructions will be delivered to dump the memory from the phone onto the computer.  A downside of this method is that the format of this data will be binary.  However, deleted data can be recovered with this process.  When using the JTAG method, you connect to test access ports on the mobile device, then send commands to the processor to transfer the raw data stored on memory chips.  This technique is a favorite among digital forensics professionals because you can gain access to the mobile devices memory without risking the integrity of the device.  However, this procedure takes more time and requires an extensive knowledge of JTAG for different models of mobile devices as well as an approach to arrange the binary data retrieved.  Invasive mobile forensics data extraction methods are more time consuming and more complex.  These methods are ideal when a device has been severely damaged and is no longer functional.  In this case the flash memory chips from the device will have to be removed and a forensic image taken.  The types of invasive mobile forensics data collection techniques are chip off and micro read.  The chip off method refers to the data being retrieved directly from the mobile device’s memory chip.  First the chip would be detached from the mobile device and inserted into a chip reader or, if applicable, a second phone could be used to read the data from the chip.  Cloning the chip from the mobile device may also be necessary if it is not severely damaged. The chip off technique is difficult because there are different types of memory chips used in mobile manufacturing.  There is a lot of training that an examiner must complete in order to apply this method in any investigation.  Additionally, special hardware may be required to work with the memory chips from the devices.  The other invasive technique used for data extraction on mobile devices is micro read.  The micro read method requires the usage of electron microscope to analyze data seen physically on the memory chip.  This technique requires the maximum level proficiency from the mobile forensic expert.

>>>>

\section{Mobile Forensics}
\subsection{Introduction to Mobile Forensics}
Mobile device forensics is a subcategory of digital forensics.  It involves the recovery of digital evidence or data from any mobile device.  These mobile devices include mobile phones, smart phones, tablets, personal digital assistants (PDAs), and Global Positioning Systems (GPS).  Another type of mobile device is wearable technology, for example an Apple watch or Fitbit fitness tracker.  These types of mobile devices can store several types of private information.  This information consists of call history, contact list, incoming and outgoing text messages, pictures, videos, audio files and voicemail messages. With the usage of smart phones, the data set also comprises internet browsing history, previous Wi-Fi connections, calendars, analytical information, geolocation data, and documents.  Some people have been known to save notes with account user credentials, passwords, and passcodes in their mobile devices as well.  All this information could be useful in an investigation.  Given the wide variety of mobile devices, there are different ways each device version must be handled.  The differentiation of the data on the device depends on the cell phone generation, the device brand, and the version of the specific model.  As a part of digital forensics investigations, there are specific tools that should be used to ensure the integrity of the mobile devices during transport from one place to the next.  Additionally, there are various data collecting techniques used for data analysis and preservation \cite{duc_2015}.

\subsection{Progression of the Cellular Networks}
Cellular networks and technologies have changed massively since their arrival in the late 1970s.  The first-generation mobile networks started in Japan in 1979 before other countries like the US and United Kingdom.  This technology was known as the advanced mobile phone system (AMPS).  The AMPS used frequency division multiple modulation and only had a capacity of 30 kilohertz and 2.4 kbps of speed.  The 1g network only allowed voice calls and had limited protection against hackers.  The second-generation network started around 1991.  This system was based on digital signals from the global system for mobile communications (GSM).  The 2g network offered bandwidths of 30 kHz to 200 kHz.  The users could also send SMS/MMS messages up to 64 kb.  Soon after, 2.5g was established and enabled data rates up to 144 kbps.  These rates allowed for browsing the web and sending   and receiving email messages. In the year 2000, the third-generation network was introduced in Europe as UMTS and CDMA in the US.  The introduction of 3G became more about social connections and supported higher speeds.  The 3G upgrade also enabled video calls, sharing files, watching television, and more surfing the internet and now for playing online games.  The fourth-generation network era brought about the usage of smartphones and other handheld devices.  These devices have speeds from 10 Mbps and 1 Gbps.  Fourth generation was the first cellular network generation to use long term evolution (LTE), which offers users better voice quality, less buffering time, instant messaging and social media, faster downloads, and quality streaming.    The 4G cellular network is also the first IP based mobile network with the technology to accommodate the quality of service and requirements including broadband access, multimedia messaging services (MMS), video chat, mobile TV, HDTV content, and digital video broadcasting.  As the 4g network tries to keep up with the demand of augmented reality and development of autonomous vehicles, the 5g revolution has begun.  The 5th generation network is being created to enhance the growing rate of new mobile applications and standards.  The output of 5G devices is 1000 times faster than 4G at 10 Gbps and will support the number of IoT (internet of things) sensors and devices \cite{keenan_2020}.

@@

Data collecting and preservation for mobile device forensics is getting more complicated as the devices advance.  Mobile devices are used for both personal and professional purposes, so the data contained reflects that.  To recover the digital evidence and other relevant data, you must follow the guidelines for preservation and examination.  All the methods and techniques must be forensically sound, meaning all actions and processes used must respect international guidelines and avoid the alteration of the original devices.  For mobile device forensics, digital forensics in general as well, evidence should always be preserved in process in a manner that allows it to be used in a court of law.  When moving mobile devices from the scene, there are two things to consider: lock activation and network/cellular connection.  Preventing network connection is important to ensure that the mobile device is not changed remotely.  The mobile device can be isolated by placing the device on airplane mode to disable Wi-Fi and hotspots.  It is also important to make sure the device has a full charge to avoid shutdown.  During storage and transportation, Faraday bags are a great tool to use.  A simple Faraday bag will isolate the device from receiving radio signals while keeping the device charged with an emergency battery.  All original cords and chargers should be removed from the scene and secluded away from the device.  Faraday cases are made from silver, copper, or nickel.  They have RoHS (Restriction of Hazardous Substances) double layer conductors, which are best for blocking EMF (electromagnetic fields) signals.  Another tool that could be used for isolation during transportation is a jammer.  A jammer sends radio waves with the same frequency as mobile phones, which causes interference between communications and paralyzes every phone in the range of the jammer.  In some countries jammers are illegal and are only allowed to be used by law enforcement.  There are many data collecting techniques.  Your plan of execution as a digital forensics professional, must include an outline of actions taken in sequence.  This is necessary because the process may have to be duplicated at another time.  Mobile forensics data collecting methods can be non-invasive and invasive.  As a good rule of thumb, always start with non-invasive techniques because “they tend to endanger the device is integrity to a lesser degree” \cite{Kostadinov_2019}.  Types of noninvasive mobile forensics data extraction methods include manual extraction, logical extraction, hex dump, and JTAG.  When arriving at the crime scene some devices may already be unlocked.  If this is the case, then manual extraction is possible.  Manual extraction is when you browse through the mobile device’s data by using the touch screen or keypad.  Ideally during manual extraction, you would document and take photos of the device while you're browsing.  Logical extraction involves connecting the mobile device to a forensically sound workstation.  The connection is established by using USB cords, Bluetooth, or RJ-45 cables.  Command requests for data retrieval from the computer to the mobile device.  With this method, data could be added to the mobile device altering the integrity.  The hexdump process is performed when the mobile device is connected to a forensically sound workstation, then using a bootloader a set of instructions will be delivered to dump the memory from the phone onto the computer.  A downside of this method is that the format of this data will be binary.  However, deleted data can be recovered with this process.  When using the JTAG method, you connect to test access ports on the mobile device, then send commands to the processor to transfer the raw data stored on memory chips.  This technique is a favorite among digital forensics professionals because you can gain access to the mobile devices memory without risking the integrity of the device.  However, this procedure takes more time and requires an extensive knowledge of JTAG for different models of mobile devices as well as an approach to arrange the binary data retrieved.  Invasive mobile forensics data extraction methods are more time consuming and more complex.  These methods are ideal when a device has been severely damaged and is no longer functional.  In this case the flash memory chips from the device will have to be removed and a forensic image taken.  The types of invasive mobile forensics data collection techniques are chip off and micro read.  The chip off method refers to the data being retrieved directly from the mobile device’s memory chip.  First the chip would be detached from the mobile device and inserted into a chip reader or, if applicable, a second phone could be used to read the data from the chip.  Cloning the chip from the mobile device may also be necessary if it is not severely damaged. The chip off technique is difficult because there are different types of memory chips used in mobile manufacturing.  There is a lot of training that an examiner must complete in order to apply this method in any investigation.  Additionally, special hardware may be required to work with the memory chips from the devices.  The other invasive technique used for data extraction on mobile devices is micro read.  The micro read method requires the usage of electron microscope to analyze data seen physically on the memory chip.  This technique requires the maximum level proficiency from the mobile forensic expert.
